---
title: "DS 202 Rollinger Final Project"
output: 
  github_document:
    number_section: FALSE
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = TRUE)


knit_hooks$set( fig.cap = function(before, options, envir) {
  if ( before ) {
    chr <- sprintf("<p><small><strong><a name='fig:%s' value = %d>Figure %d</a></strong>: %s</small></p>", options$label, options$fig.num, options$fig.num, options$fig.cap)
    chr <- sprintf("<p><small><strong><a name='fig:%s'>%s</a></strong>: %s</small></p>", options$label, options$label,  options$fig.cap)
    
    return(chr)
  }
})


chunkref <- local({
  function(chunklabel) {
    sprintf('[%s](#%s)', chunklabel, paste0("fig:",chunklabel) )
  }  
})


```

# Analysis of Iowa Counties

Authors: Ethan Rollinger

## Abstract

This final project is examining the data set "county_complete", which is a data set that that holds information about each county in the entire country of the United States. The data is from the U.S. Census Bureau's American Community Survey, which provides a large portion of county-level characteristics. The main focus of this final project will be looking at the population growth among the counties from the information provided (2000 to 2017), along with looking at the racial, ethnic and veteran composition for each county in Iowa. This in turn will provide information and knowledge of the key demographic and social factors and their variance geographically within the state. The motivation for completing this project is to understand and explore the general demographic of each county in Iowa and see what ideas and insights can come from this for potential future studies and analyses. The main findings for this project are:

- 1. Counties with the highest population growth between 2000 and 2017 were primarily located Central and Eastern, usually around larger urban areas and cities
- 2. The average racial and ethnic demographic had a very large range of variability, with the more diverse populations being located in large metropolitan areas such as Des Moines and Iowa City
- 3. Counties that had the highest percentage of veterans were more rural, and also consisted of slower population growth and on average older communities. 


# Introduction

This topic of this project is the exploration of three key aspects of county-level demographics in Iowa: population growth between 2000 and 2017, the average racial and ethnic composition of each county, and the percentage of county residents who are military veterans. These factors are important for understanding how local communities are changing over time and how population shifts, diversity, and veteran representation vary by region. Although these topics are often studied at the national level, a smaller county-based analysis can show patterns and potential correlations that may be unseen or overlooked. 

This analysis is relevant because demographic trends such as population growth, ethnicity rates and veteran composition can influence decision-making on the government level. Some of these decisions can be in areas such as healthcare access, infrastructure planning, education funding, and social services. Understanding where populations are growing, which areas are becoming more diverse, and where veteran communities are concentrated can help influence policy making and action.

This work is based on the county_complete data set, which compiles data from the U.S. Census Bureau and the American Community Survey. The data set includes multiple years of observations across all U.S. counties, allowing a variety of comparisons to be made. For this project, a subset of this data will be analyzed by looking at only Iowa. This is done for better visuals of graphs. 

In Section 1 (Results 1), population growth trends from 2013 to 2017 will be shown and which counties and regions experienced the most change. Section 2 (Results 2) examines the racial and ethnic makeup of counties to inform patterns of diversity. Finally, Section 3 (Results 3) analyzes veteran status across counties with a larger focus on differences between urban and rural areas.


# Data Summary

```{r, echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)

options(tigris_use_cache = TRUE)
```

The above code contains all of the packages that I will be using for this project. 

```{r, echo=TRUE}
county_complete <- read.csv("https://www.openintro.org/data/csv/county_complete.csv")
View(county_complete)
```

The data set, as mentioned before comes from the U.S. Census Bureau and the American Community Survey and can be found from the following link: https://www.openintro.org/data/index.php?data=county_complete. As this data set has been commonly used, minimal cleaning needs to be done to this data set. It is compiled of data for all 3142 counties in the United States. It contains 188 variables, where it has information discussing population, percentage of ethnicities, education background, household income and much more. For this project, I will be focusing on the population, ethnicity demographic and veteran composition, so the following variables will be important:

- state (State)
- name (County Name)
- fips (FIPS Code)
- pop2000 (Population in 2000)
- pop2010 (Population in 2010)
- pop2011 (Population in 2011)
- pop2012 (Population in 2012)
- pop2013 (Population in 2013)
- pop2014 (Population in 2014)
- pop2015 (Population in 2015)
- pop2016 (Population in 2016)
- pop2017 (Population in 2017)
- white_2010 (Percent of population that is white in 2010)
- black_2010 (Percent of population that is black in 2010)
- black_2017 (Percent of population that is black in 2017)
- native_2010 (Percent of population that is a Native American in 2010)
- native_2017 (Percent of population that is a Native American in 2017)
- asian_2010 (Percent of population that is a Asian in 2010)
- asian_2017 (Percent of population that is a Asian in 2017)
- pac_isl_2010 (Percent of population that is Hawaii or Pacific Islander in 2010)
- pac_isl_2017 (Percent of population that is Hawaii or Pacific Islander in 2017)
- hispanic_2010 (Percent of population that is Hispanic in 2010)
- hispanic_2017 (Percent of population that is Hispanic in 2017)
- veterans_2010 (Percent of population that are veterans from 2006-2010)
- veterans_2017 (Percent of population that are veterans in 2017)

```{r, echo=TRUE}

county_selected <- county_complete %>%
  select(state,name,fips,pop2000,pop2010,pop2011,pop2012,pop2013,pop2014,pop2015,pop2016,pop2017,white_2010,black_2010,black_2017,native_2010,native_2017,asian_2010,asian_2017,pac_isl_2010,pac_isl_2017,other_single_race_2017,hispanic_2010,hispanic_2017,veterans_2010,veterans_2017)

View(county_selected)

```

The above code cleaned the data set and got rid of the unneeded variables for this project. Using the above variable list, each variable was selected to create the new data set "county_selected". This was done to simplify and condense the important data and to create a more readable data set. 

```{r, echo=TRUE}
iowa_counties <-county_selected %>% filter(state == "Iowa")
```

The above code cleaned the data set even further to create a new data set called "iowa_counties" in order to create a data set that only focuses on data from Iowa counties. This will be the data set used for the rest of the project. 

# Results

## Results 1: Population Growth of Iowa Counties

This section analyzes the population changes across U.S. counties from 2000 to 2017. In depth population trends will be looked at, along with different visuals that explore the comparisons between counties throughout the given years. 

```{r, echo=TRUE}
iowa_counties <- iowa_counties %>%
  mutate(pop_change_2000_2017 = pop2017 - pop2000)

View (iowa_counties)
```

This line of code inserted a new column that shows the total population change between the range limits of the years (2000 and 2017). 

### Population Growth from 2000 to 2017
```{r, echo=TRUE}
ggplot(iowa_counties, aes(x = pop_change_2000_2017)) +
  geom_histogram(fill = "lightblue", color = "black", binwidth = 500) +
  labs(
    title = "Density of Population Change (2000-2017)",
    x = "Population Change (2000-2017)",
    y = " US Counties"
  ) + scale_x_continuous(limits = c(-5000, 20000))

```

The plot consists of a histogram showing the distribution of population change from 2000 to 2017. The graph shows that most counties experienced zero or negative population growth, with the bell curve being centered around the -2000 range. However, there are four outliers on the higher end of the population change spectrum, indicating a few counties with unusually high growth, creating some curiosity and potential followup as to why there is such a large increase in population growth. 

### Population Growth from Each Year
```{r, echo=TRUE, message=FALSE, warning=FALSE}
pop_data_long <- iowa_counties %>%
  gather(key = "year", value = "population", pop2000:pop2017)

ggplot(pop_data_long, aes(x = year, y = population, group = name, color = name)) +
  geom_line(alpha = 0.5, size = 1) +
  labs(title = "Population Changes in U.S. Counties (2000-2017)",
       x = "Year", y = "Population") + scale_x_discrete(labels = function(x) gsub("pop", "", x)) +
  theme_minimal() +
  theme(legend.position = "none")

```

This plot shows population trends for each county from 2000 to 2017, with each line representing a different county. The population generally is flat over time, but some hold little growth in more urban counties.The outliers can be found in more urban areas (Polk County) as mentioned before, which suggests for more exploration into local demographic factors.

### Comparing Population Across Counties from 2017

```{r, echo=TRUE, message=FALSE}
iowa_map <- tigris::counties(state = "IA", cb = TRUE, class = "sf")

pop_data <- iowa_counties %>%
  mutate(name = gsub(" County", "", name)) %>%
  select(name, pop2017) %>%
  rename(population = pop2017)

pop_data_long <- iowa_counties %>%
  mutate(name = gsub(" County", "", name)) %>%
  pivot_longer(cols = starts_with("pop"), 
               names_to = "year", 
               values_to = "population") %>%
  mutate(year = gsub("pop", "", year))


iowa_map$name <- gsub(" County", "", iowa_map$NAMELSAD)


iowa_merged <- iowa_map %>%
  left_join(pop_data, by = "name")

ggplot(iowa_merged) +
  geom_sf(aes(fill = population), color = "white") +
  scale_fill_viridis_c(option = "viridis") +
  labs(title = "Iowa County Population (2017)", fill = "Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),  
    axis.ticks = element_blank(),  
    axis.title = element_blank()) 


```

This plot visualizes the population in 2017 for each county. The population from 2017 is generally higher in counties with large cities, such as Polk, Linn and Johnston County. Some counties exhibit very low population, mainly found in the rual parts of the State, which helps explain why there population is so low. 

Errors I ran into in this section were creating the Iowa map. Slowly walking myself through the correct steps to make the map and have the data correctly merged into it took some time. I spent a great deal of time going back and forth from stackoverflow and chat gpt to correctly have everything line up. I also had some struggles with the formatting of the graphs. I wanted to make everything visually appealing, so slowly tweaking the theme functions of each graph also took some time. 

## Results 2: Racial and Ethnic Composition

This section analyzes the ethnic composition of Iowa counties, comparing data from 2010 and 2017. The analysis will focus on the following racial groups: African American, Hispanic, Asian, Native American and Pacific Islander.

### Racial Composition: African Americans
```{r, echo=TRUE}

ethnicity_pct_black <- iowa_counties %>%
  mutate(name = gsub(" County", "", name))

ethnicity_long_black <- ethnicity_pct_black %>%
  pivot_longer(
    cols = c(black_2010, black_2017),
    names_to = c("race", "year"),
    names_sep = "_",
    values_to = "percent"
  )

iowa_map_long_black <- iowa_map %>%
  left_join(ethnicity_long_black, by = "name")

ggplot(iowa_map_long_black) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(race ~ year) +
  labs(title = "Iowa County African American Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),  
    axis.ticks = element_blank(),  
    axis.title = element_blank()  
  )

```

This map illustrates the percentage of African American population in Iowa counties, separated by year (2010 and 2017). The main finding reveals that African American population percentages have stayed relatively stable in most counties, with only slight increases in urban counties between 2010 and 2017. Outliers are seen in counties with larger urban populations, where the percentage of African residents is more significant compared to rural areas, where the percentage is lower.

### Racial Compsoition: Hispanics
```{r, echo=TRUE}
ethnicity_pct_hispanic <- iowa_counties %>%
  mutate(name = gsub(" County", "", name))

ethnicity_long_hispanic <- ethnicity_pct_hispanic %>%
  pivot_longer(
    cols = c(hispanic_2010, hispanic_2017),
    names_to = c("race", "year"),
    names_sep = "_",
    values_to = "percent"
  )

iowa_map_long_hispanic <- iowa_map %>%
  left_join(ethnicity_long_hispanic, by = "name")

ggplot(iowa_map_long_hispanic) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(race ~ year) +
  labs(title = "Iowa County Hispanic Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),  
    axis.title = element_blank()   
  )

```

The map visualizes the percentage of Hispanic population in Iowa counties, facet-wrapped by year (2010 and 2017).The main finding shows some variation in Hispanic population percentages across counties, with some counties experiencing an increase from 2010 to 2017. Outliers include a few counties with exceptionally high percentages, such as those near large urban centers, while many counties show very low Hispanic populations.

### Racial Composition: Asians
```{r, echo=TRUE}

ethnicity_pct_asian <- iowa_counties %>%
  mutate(name = gsub(" County", "", name))

ethnicity_long_asian <- ethnicity_pct_asian %>%
  pivot_longer(
    cols = c(asian_2010, asian_2017),
    names_to = c("race", "year"),
    names_sep = "_",
    values_to = "percent"
  )

iowa_map_long_asian <- iowa_map %>%
  left_join(ethnicity_long_asian, by = "name")

ggplot(iowa_map_long_asian) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(race ~ year) +
  labs(title = "Iowa County Asian Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    axis.title = element_blank()  
  )

```

The map shows the percentage of Asian population in Iowa counties, divided by year (2010 and 2017). The main finding indicates that some counties have seen a noticeable increase in their Asian population from 2010 to 2017, particularly in more metropolitan areas. Outliers include counties with substantial growth in Asian populations, likely due to international migration or relocation, while many rural counties report low percentages of Asian residents.

### Racial Compsoition: Native Americans
```{r, echo=TRUE}

ethnicity_pct_na <- iowa_counties %>%
  mutate(name = gsub(" County", "", name))

ethnicity_long_na <- ethnicity_pct_na %>%
  pivot_longer(
    cols = c(native_2010, native_2017),
    names_to = c("race", "year"),
    names_sep = "_",
    values_to = "percent"
  )

iowa_map_long_na <- iowa_map %>%
  left_join(ethnicity_long_na, by = "name")

ggplot(iowa_map_long_na) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(race ~ year) +
  labs(title = "Iowa County Native American Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(), 
    axis.title = element_blank() 
  )

```

The map visualizes the percentage of Native American population in Iowa counties, facet-wrapped by year (2010 and 2017). The main finding shows that the Native American population is relatively small in most counties, with only a few counties showing noticeable changes from 2010 to 2017. Outliers include Tama County, which has more significant Native American populations due to there being a reservation in this county. 

### Racial Composition: Pacific Islanders
```{r, echo=TRUE}
colnames(iowa_counties) <- gsub("pac_isl", "pacisl", colnames(iowa_counties))

ethnicity_pct_pacisl <- iowa_counties %>%
  mutate(name = gsub(" County", "", name))

ethnicity_long_pacisl <- ethnicity_pct_pacisl %>%
  pivot_longer(
    cols = c(pacisl_2010, pacisl_2017),
    names_to = c("race", "year"),
    names_sep = "_",
    values_to = "percent"
  )

iowa_map_long_pacisl <- iowa_map %>%
  left_join(ethnicity_long_pacisl, by = "name")

ggplot(iowa_map_long_pacisl) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(race ~ year) +
  labs(title = "Iowa County Pacific Islander Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),  
    axis.ticks = element_blank(),
    axis.title = element_blank()  
  )

```

This map highlights the percentage of Pacific Islander population in Iowa counties, facet-wrapped by year (2010 and 2017). The main finding shows that the Pacific Islander population remains quite small across most counties, with minimal changes between 2010 and 2017, along with many missing data from 2010. Outliers include counties with small, yet significant, Pacific Islander populations, particularly in areas with local immigrant communities, while other counties have negligible percentages.

I ran into a lot of errors with this section, mainly with the formatting of the maps. I originally had all of the maps shown in one visual, but the color scale was not very even, as it only produces different colors for the hispanic populations as they are the majority of the minority groups. To combat this problem, I switched to making them all their own maps to better shown the population differences. I also had trouble with creating new data sets and function and merging them all into the map. 

## Results 3: Veteran Status

In this section, we examine the veteran status across counties for both 2010 (data from 2006-2010) and 2017. This is useful for understanding how veteran populations are distributed and how this has changed over time in Iowa. 

### Veteran Percentage in 2010 and 2017
```{r, echo=TRUE}
iowa_counties <- iowa_counties %>%
  rename(veteran_2010 = veterans_2010)

iowa_counties <- iowa_counties %>%
  mutate(veterans_2010 = 100 * veteran_2010 / pop2010)
```

There was an error in the dataset, as veterans_2010 listed the total number of veterans living in each county, where veterans_2017 listed the percentage of the population that are considered veterans. The above code creates a new column (veteranspct_2010) which will be used instead of veterans_2010 to create the following graph. 

```{r, echo=TRUE}
veteran_pct <- iowa_counties %>%
  mutate(
    veterans_2010 = 100 * veteran_2010 / pop2010,
    name = gsub(" County", "", name)
  )

veteran_long <- veteran_pct %>%
  pivot_longer(
    cols = c(veterans_2010, veterans_2017),
    names_to = c("group", "year"),
    names_sep = "_",
    values_to = "percent"
  )


iowa_map_long_veteran <- iowa_map %>%
  left_join(veteran_long, by = "name")

ggplot(iowa_map_long_veteran) +
  geom_sf(aes(fill = percent), color = "white") +
  scale_fill_viridis_c(option = "plasma", na.value = "gray") +
  facet_grid(group ~ year) +
  labs(title = "Iowa County Veteran Population by Year", fill = "% of Population") +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )

```

The map displays the percentage of veterans in Iowa counties, using a facet grid to compare 2010 and 2017 values. The main finding shows a slight decline in veteran population percentages in many counties from 2010 to 2017, likely reflecting aging demographics. Some rural counties exhibit notably higher veteran percentages, while others—especially urban centers—show lower values, suggesting differing patterns of settlement or veteran services availability.

This results section was relatively easy, as it followed a similar style to the result #2. One trouble I did have was the veteran_2010 and vetern_2017 differences. One variable was in percentages while the other was total population, so I had some initial trouble in the maps. Once I figured this problem out, the rest went smoothly. 

# Conclusions

This analysis explored population growth, ethnic composition, and veteran status across Iowa counties from 2000 to 2017. The results showed that while it is true that not all counties experienced the same levels, some did see population growth over that time. Most, however, saw little to no growth, and some even regressed. Ethnic composition also shifted over time, with notable changes in the percentage of African American, Hispanic, and Native American populations from 2010 to 2017, reflecting broader demographic trends in the country. Additionally, veteran status data revealed a decline in the percentage of veterans in many counties, suggesting changing dynamics in this population group.Veterans were also more spread into rural areas rather than urban areas. A limitation of this analysis is the lack of detailed socio-economic variables that could further explain these patterns, along with not as current of data. With more time or team members, it would have been beneficial to explore the other variables in this data set, such as education and household income, and see if this ties into the demographics discussed. 

## References

- https://www.openintro.org/data/index.php?data=county_complete
- https://stackoverflow.com/questions
- https://chatgpt.com/
